The Talking Heads Experiment

### 2015-04-10 chap4/chap4.tex
quotation marks for chapters 8 to 10

Consider the scene in \figref{geom}. Assuming that the agent
already has a well-developed set of categories, he could use 
the category [GRAY-0.75,1.0] (very dark) to distinguish 
shape 3 from the others. On the other hand, if shape 2 is 
chosen as topic, the grayscale will not be enough because
shape 2 has the same grayscale as shape 1. Maybe a 
combination of categories can be chosen, like [VPOS 0.0--05]
(lower), and [HPOS 0.0–0.5] (left). Indeed shape 2
is lower in the scene, as opposed to
shape 1 and shape 3, and it is more to the left compared 
to shape 0 and 2.

@@

A new subcategoriser is constructed by taking a
categoriser node in the tree and dividing its range into 
two new subranges and thus two new subcategorisers.
For example, if there is a 
categoriser [HPOS 0.0–0.5], which triggers when the 
object is in the left most half of a scene, i.e. with HPOS
within $[0.0,0.5]$, then two 
subcategories are created by dividing $[0.0,0.5]$
into two halves, one for the 
range $[0.0,0.25]$ ([HPOS 0.0–0.25] or totally left) 
and one for the range $[0.25,0.5]$ ([HPOS 0.25--0.5] or mid-left). 
A new categoriser is added to the tree for each of these
halves.

>>>>

Consider the scene in \figref{geom}. Assuming that the agent
already has a well-developed set of categories, he could use 
the category [GRAY-0.75,1.0] (very dark) to distinguish 
shape 3 from the others. On the other hand, if shape 2 is 
chosen as topic, the grayscale will not be enough because
shape 2 has the same grayscale as shape 1. Maybe a 
combination of categories can be chosen, like [VPOS 0.0–05]
(lower), and [HPOS 0.0–0.5] (left). Indeed shape 2
is lower in the scene, as opposed to
shape 1 and shape 3, and it is more to the left compared 
to shape 0 and 2.

@@

A new subcategoriser is constructed by taking a
categoriser node in the tree and dividing its range into 
two new subranges and thus two new subcategorisers.
For example, if there is a 
categoriser [HPOS 0.0–0.5], which triggers when the 
object is in the left most half of a scene, i.e. with HPOS
within $[0.0,0.5]$, then two 
subcategories are created by dividing $[0.0,0.5]$
into two halves, one for the 
range $[0.0,0.25]$ ([HPOS 0.0–0.25] or totally left) 
and one for the range $[0.25,0.5]$ ([HPOS 0.25–0.5] or mid-left). 
A new categoriser is added to the tree for each of these
halves.

### 2015-04-10 chap7/chap7.tex
quotation marks for chapters 8 to 10

I now show a series of example games taken from 
an experiment in which two situated embodied agents,
{\bfshape a1} and {\bfshape a2}, play grounded language games based on 
coloured figures pasted on the white board in front
of them. The following channels are available to the agents 
in the experiments in this section: 
HPOS (the horizontal position of the midpoint 
of the segment's bounding box), 
VPOS (the vertical position of the midpoint of the segment's 
bounding box), HEIGHT (the height of the bounding box), 
WIDTH (the width of the bounding box), AREA
(the area of the segment, calculated
by counting the number of pixels that belong to it), 
R (the average redness of the pixels in the segment), 
G (the average greenness of the pixels in the segment), 
B (the average blueness of the pixels in the segment). 
These `colour' channels are not to be confused with the 
human opponent colour channels so the distinctions 
that form of them are not directly perceivable by 
human observers.

@@

Such an ideal language has been a dream of many 
philosophers, including Descartes and Leibniz, but 
the investigations reported here show clearly that 
it is not attainable.\footnote{The history of this search for such an ideal language
is discussed by \cite{Eco:1997}.}
A language must be open to the 
expression of new meanings because the communicative 
objectives and the environments which are the subject
of communication keep changing. Hence 
synonymy (incoherence along the MF dimension) is unavoidable
because words are created by some agents not knowing that 
words already exist in the population for the same meaning. 
In addition, one agent 
may `wrongly' infer that a certain word has a 
particular meaning due to perceptual anomalies, even though
the same word had already another meaning in the
lexicon. Word-meaning pairs that thus
arise may start to propagate 
in the population and actually supersede the `original' 
word-meaning pairs.

>>>>

I now show a series of example games taken from 
an experiment in which two situated embodied agents,
{\bfshape a1} and {\bfshape a2}, play grounded language games based on 
coloured figures pasted on the white board in front
of them. The following channels are available to the agents 
in the experiments in this section: 
HPOS (the horizontal position of the midpoint 
of the segment's bounding box), 
VPOS (the vertical position of the midpoint of the segment's 
bounding box), HEIGHT (the height of the bounding box), 
WIDTH (the width of the bounding box), AREA
(the area of the segment, calculated
by counting the number of pixels that belong to it), 
R (the average redness of the pixels in the segment), 
G (the average greenness of the pixels in the segment), 
B (the average blueness of the pixels in the segment). 
These ``colour'' channels are not to be confused with the 
human opponent colour channels so the distinctions 
that form of them are not directly perceivable by 
human observers.

@@

Such an ideal language has been a dream of many 
philosophers, including Descartes and Leibniz, but 
the investigations reported here show clearly that 
it is not attainable.\footnote{The history of this search for such an ideal language
is discussed by \cite{Eco:1997}.}
A language must be open to the 
expression of new meanings because the communicative 
objectives and the environments which are the subject
of communication keep changing. Hence 
synonymy (incoherence along the MF dimension) is unavoidable
because words are created by some agents not knowing that 
words already exist in the population for the same meaning. 
In addition, one agent 
may ``wrongly'' infer that a certain word has a 
particular meaning due to perceptual anomalies, even though
the same word had already another meaning in the
lexicon. Word-meaning pairs that thus
arise may start to propagate 
in the population and actually supersede the ``original'' 
word-meaning pairs.

### 2015-04-10 chap8/chap8.tex
quotation marks for chapters 8 to 10

\begin{figure}[htbp]
  \centerline{\includegraphics[width=0.7\textwidth]{chap8/figures/hu-obrist+ls}}
\caption{ \label{fig:faces} 
Hans-Ulrich Obrist and Luc Steels during the opening of the Laboratorium Exhibition in Antwerp, summer 1999. }
\end{figure}
In addition to the installation at the exhibition in Antwerp itself, there were two additional 
external sites for the Talking Heads experiment operating within the same time frame: The Paris site at the Sony Computer 
Science Laboratory in Paris and the Brussels site at the Free University (Vrije Universiteit) of Brussels.

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=0.9\textwidth]{chap8/figures/layout}}
\caption{\label{fig:layout} 
Layout of the 'Laboratory for Cognitive Robotics and Teleporation" at the Laboratorium exhibition in Antwerp.}
\end{figure}

\begin{figure}[htbp]
 \centerline{\includegraphics[width=.35\textwidth]{chap8/figures/zoom-in-head}}
\caption{ \label{fig:laboratorium} 
A single Talking Head at the Paris site. In the background we see (bottom) the display with an
image that the agent picked up, (on top of that) a screen displaying the processed image and the discrimination trees, 
and to the right the loudspeaker broadcasting the speech output of the agent.}
\end{figure}

A typical example of the kind of images that the camera's picked up are shown in \figref{fig:images}. 
To play a game, the agent had to find a sufficiently delineated group of objects and each object had 
to exceed a minimal size. Initially quite complex configurations were put on the white board but this made it very difficult for agents to find a coherent group and to develop good concepts for referring to the topic. The right 
image in \figref{fig:images} provides clear examples of up, down, left and right, different colors, and 
also opportunities to use multiple words (such as ``red bottom").

\begin{figure}[htbp]
 \centerline{\includegraphics[width=.40\textwidth]{chap8/figures/image-1}
\includegraphics[width=.40\textwidth]{chap8/figures/image-2}}
\caption{\label{fig:images} 
Examples of images picked up by the camera's. The chosen topic was signalled by drawing a bounding box around it. 
During experimentation, it was found that configurations such as in the right image lead to more stable performance.} 
\end{figure}

@@

\begin{figure}[htbp]
 \centerline{\includegraphics[width=.80\textwidth]{chap8/figures/xagadu-speaker}}
\centerline{\includegraphics[width=.80\textwidth]{chap8/figures/xagadu-hearer}}
\caption{\label{fig:interaction} 
Top: Source image, conceptualisation, and word choice of the speaker. Bottom: Source image, parsing and interpretation 
of the hearer.}
\end{figure}

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.85\textwidth]{chap8/figures/th-website}}
\caption{\label{fig:thwebsite} 
The Talking Heads website main interface. Users could create and manage their own agents and inspect what was happening 
at each site. Here 12 agents located at the Brussels site are listed.}
\end{figure}

An example of how the web interface displayed a single game is shown in \figref{fig:thwebsite-agent}. 
There are two agents: Antonusius is the speaker and Zelebot is the hearer. 
The green object has been chosen as topic and correctly 
recognised by the hearer using the word ``kazozo". The meaning of the word was not visible through the interface because we 
wanted users to learn themselves the language of the agents.

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.85\textwidth]{chap8/figures/zelebot}}
\caption{\label{fig:thwebsite-agent} 
Example of an interaction that took part in Antwerp. This game was a success. The agents recognised the same
set of objects and the meaning of ``kazozo" was effective in finding the same topic. 
}
\end{figure}

@@

The first `Talking Heads' experiment ran for 4 months during the summer of 1999 and showed the validity of the mechanisms that were used for the agent architecture and of the interaction patterns and group dynamics of the agents. A shared lexicon and an underlying conceptual repertoire emerged, enabling successful communication by the agents about the scenes before them. In total, 400,000 grounded games were played. The population of agents rose to just under 2000, increasing steadily over the period of the experiment. Despite the many perturbations due to grounding, intermittent technical failures, a continuous influx of new agents entering the population, and unpredictable human interaction, the lexicon was maintained throughout this period.

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.80\textwidth]{chap8/figures/allsites}}
\caption{\label{fig:allsites} 
The y-axis shows the average communicative success of agents at each of the three different sites for a total 
of 190.000 games shown on the y-axis. 
}
\end{figure}

@@

\begin{figure}[htbp]
 \centerline{\includegraphics[width=.80\textwidth]{chap8/figures/words-used}}
\caption{\label{fig:words-used} 
Distribution of word use. There are only a few words that are dominant. Many words are 
short-lived, either because the circumstances 
in which they fit are rare or because they do not get settled in the population. Moreover sometimes users create agents 
which they do not keep rescheduling for new games. 
}
\end{figure}

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.80\textwidth,height= 6cm]{chap8/figures/wogglesplat}}
\caption{\label{fig:woggle} 
Meaning-Form diagram: Different words for expressing the meaning 'left', i.e. 
horizontal position is less than 0.25 (scaled). 
New words come up all the time but there is a clear winner-take-all effect with "wogglesplat" winning. 
}
\end{figure}

\figref{fig:droite} shows another example of the semiotic dynamics in the experiment, this time looking at all the 
meanings for a particular word, namely ``droite" (meaning `to the right' in French). This word clearly has been introduced
by a human user and the dominating meaning progressively becomes a region (between 0.75 and 1.0) on the horizontal position, 
as could be expected. On the way we see some confusion, Specifically there must be objects that appear both on the right 
and up (vertical position between 0.5 and 1.0).

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.80\textwidth,height=6cm]{chap8/figures/droite}}
\caption{\label{fig:droite} 
Form-Meaning diagram: Different meanings associated with the word ``droite". The dominant meaning is 
in line with the human use of the word, namely horizontal position (HPOS) to the right of the image.}
\end{figure}

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.70\textwidth, height=6cm]{chap8/figures/bozopite}}
\caption{\label{fig:bozopite} 
Form-Meaning diagram: Different meanings associated with the word ``bozopite". 
}
\end{figure}

It was also possible that different meanings which were highly compatible were maintained in the population. 
An example is shown in the form-meaning diagram in \figref{fig:down}, which plots the average 
frequency of meanings for the word ``down". The meanings are all concepts on the vertical position channel 
VPOS, but they carve out smaller and smaller regions.

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.65\textwidth]{chap8/figures/down}}
\caption{\label{fig:down} 
Form-Meaning diagram: Different meanings associated with the word "down". There are in particular two meanings. 
one is a region on the red-channel and another one a region on the vertical position channel. 
}
\end{figure}

@@

Numerous talks were given and various papers published in scientific fora. There were also various talks within 
the art context\footnote{Examples are a 'gallery talk' as Salon 3 at the Elephant and Castle Centre
in London on 2 december 1998, organised by Hans Ulrich Obrist and Molly Nesbit, and a presentation together with 
Hans-Ulrich Obrist and Rem Koolhaas in Antwerp to launch the Laboratorium book on 3 October 2001.}
The experiment received wide coverage in 
the media thanks to its public exposure as part of a major exhibition. The Sued-Deutsche Zeitung called it 
`Angels with Internet wings'. All this lead to invitations to show the experiment
also in other venues, as discussed in the next chapter. 
Regretably, the challenge just to keep the experiment in the air with the available human resources
and the pressure for going on with new experiments prevented us from doing more adequate data gathering 
and analysis. Nevertheless some analyses appeared, particularly as carried out by Frederic Kaplan\cite{Kaplan:2001}.

@@

The experiment ended with the following `tongue-in-cheek' post by Angus McIntyre on the Forum: \\

>>>>

\begin{figure}[htbp]
  \centerline{\includegraphics[width=0.7\textwidth]{chap8/figures/hu-obrist+ls.pdf}}
\caption{ \label{fig:faces} 
Hans-Ulrich Obrist and Luc Steels during the opening of the Laboratorium Exhibition in Antwerp, summer 1999. }
\end{figure}
In addition to the installation at the exhibition in Antwerp itself, there were two additional 
external sites for the Talking Heads experiment operating within the same time frame: The Paris site at the Sony Computer 
Science Laboratory in Paris and the Brussels site at the Free University (Vrije Universiteit) of Brussels.

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=0.9\textwidth]{chap8/figures/layout.pdf}}
\caption{\label{fig:layout} 
Layout of the 'Laboratory for Cognitive Robotics and Teleporation" at the Laboratorium exhibition in Antwerp.}
\end{figure}

\begin{figure}[htbp]
 \centerline{\includegraphics[width=.35\textwidth]{chap8/figures/zoom-in-head.pdf}}
\caption{ \label{fig:laboratorium} 
A single Talking Head at the Paris site. In the background we see (bottom) the display with an
image that the agent picked up, (on top of that) a screen displaying the processed image and the discrimination trees, 
and to the right the loudspeaker broadcasting the speech output of the agent.}
\end{figure}

A typical example of the kind of images that the camera's picked up are shown in \figref{fig:images}. 
To play a game, the agent had to find a sufficiently delineated group of objects and each object had 
to exceed a minimal size. Initially quite complex configurations were put on the white board but this made it very difficult for agents to find a coherent group and to develop good concepts for referring to the topic. The right 
image in \figref{fig:images} provides clear examples of up, down, left and right, different colors, and 
also opportunities to use multiple words (such as `red bottom').

\begin{figure}[htbp]
 \centerline{\includegraphics[width=.40\textwidth]{chap8/figures/image-1.jpg}
\includegraphics[width=.40\textwidth]{chap8/figures/image-2.pdf}}
\caption{\label{fig:images} 
Examples of images picked up by the camera's. The chosen topic was signalled by drawing a bounding box around it. 
During experimentation, it was found that configurations such as in the right image lead to more stable performance.} 
\end{figure}

@@

\begin{figure}[htbp]
 \centerline{\includegraphics[width=.80\textwidth]{chap8/figures/xagadu-speaker.pdf}}
\centerline{\includegraphics[width=.80\textwidth]{chap8/figures/xagadu-hearer.pdf}}
\caption{\label{fig:interaction} 
Top: Source image, conceptualisation, and word choice of the speaker. Bottom: Source image, parsing and interpretation 
of the hearer.}
\end{figure}

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.85\textwidth]{chap8/figures/th-website.pdf}}
\caption{\label{fig:thwebsite} 
The Talking Heads website main interface. Users could create and manage their own agents and inspect what was happening 
at each site. Here 12 agents located at the Brussels site are listed.}
\end{figure}

An example of how the web interface displayed a single game is shown in \figref{fig:thwebsite-agent}. 
There are two agents: Antonusius is the speaker and Zelebot is the hearer. 
The green object has been chosen as topic and correctly 
recognised by the hearer using the word `kazozo'. The meaning of the word was not visible through the interface because we 
wanted users to learn themselves the language of the agents.

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.85\textwidth]{chap8/figures/zelebot.pdf}}
\caption{\label{fig:thwebsite-agent} 
Example of an interaction that took part in Antwerp. This game was a success. The agents recognised the same
set of objects and the meaning of `kazozo' was effective in finding the same topic. 
}
\end{figure}

@@

The first ``Talking Heads'' experiment ran for 4 months during the summer of 1999 and showed the validity of the mechanisms that were used for the agent architecture and of the interaction patterns and group dynamics of the agents. A shared lexicon and an underlying conceptual repertoire emerged, enabling successful communication by the agents about the scenes before them. In total, 400,000 grounded games were played. The population of agents rose to just under 2000, increasing steadily over the period of the experiment. Despite the many perturbations due to grounding, intermittent technical failures, a continuous influx of new agents entering the population, and unpredictable human interaction, the lexicon was maintained throughout this period.

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.80\textwidth]{chap8/figures/allsites.pdf}}
\caption{\label{fig:allsites} 
The y-axis shows the average communicative success of agents at each of the three different sites for a total 
of 190.000 games shown on the y-axis. 
}
\end{figure}

@@

\begin{figure}[htbp]
 \centerline{\includegraphics[width=.80\textwidth]{chap8/figures/words-used.pdf}}
\caption{\label{fig:words-used} 
Distribution of word use. There are only a few words that are dominant. Many words are 
short-lived, either because the circumstances 
in which they fit are rare or because they do not get settled in the population. Moreover sometimes users create agents 
which they do not keep rescheduling for new games. 
}
\end{figure}

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.80\textwidth,height= 6cm]{chap8/figures/wogglesplat.pdf}}
\caption{\label{fig:woggle} 
Meaning-Form diagram: Different words for expressing the meaning 'left', i.e. 
horizontal position is less than 0.25 (scaled). 
New words come up all the time but there is a clear winner-take-all effect with "wogglesplat" winning. 
}
\end{figure}

\figref{fig:droite} shows another example of the semiotic dynamics in the experiment, this time looking at all the 
meanings for a particular word, namely `droite' (meaning `to the right' in French). This word clearly has been introduced
by a human user and the dominating meaning progressively becomes a region (between 0.75 and 1.0) on the horizontal position, 
as could be expected. On the way we see some confusion, Specifically there must be objects that appear both on the right 
and up (vertical position between 0.5 and 1.0).

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.80\textwidth,height=6cm]{chap8/figures/droite.pdf}}
\caption{\label{fig:droite} 
Form-Meaning diagram: Different meanings associated with the word `droite'. The dominant meaning is 
in line with the human use of the word, namely horizontal position (HPOS) to the right of the image.}
\end{figure}

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.70\textwidth, height=6cm]{chap8/figures/bozopite.pdf}}
\caption{\label{fig:bozopite} 
Form-Meaning diagram: Different meanings associated with the word `bozopite'. 
}
\end{figure}

It was also possible that different meanings which were highly compatible were maintained in the population. 
An example is shown in the form-meaning diagram in \figref{fig:down}, which plots the average 
frequency of meanings for the word `down'. The meanings are all concepts on the vertical position channel 
VPOS, but they carve out smaller and smaller regions.

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.65\textwidth]{chap8/figures/down.pdf}}
\caption{\label{fig:down} 
Form-Meaning diagram: Different meanings associated with the word "down". There are in particular two meanings. 
one is a region on the red-channel and another one a region on the vertical position channel. 
}
\end{figure}

@@

Numerous talks were given and various papers published in scientific fora. There were also various talks within 
the art context\footnote{Examples are a `gallery talk' as Salon 3 at the Elephant and Castle Centre
in London on 2 december 1998, organised by Hans Ulrich Obrist and Molly Nesbit, and a presentation together with 
Hans-Ulrich Obrist and Rem Koolhaas in Antwerp to launch the Laboratorium book on 3 October 2001.}
The experiment received wide coverage in 
the media thanks to its public exposure as part of a major exhibition. The Süddeutsche Zeitung called it 
``Angels with Internet wings''. All this lead to invitations to show the experiment
also in other venues, as discussed in the next chapter. 
Regretably, the challenge just to keep the experiment in the air with the available human resources
and the pressure for going on with new experiments prevented us from doing more adequate data gathering 
and analysis. Nevertheless some analyses appeared, particularly as carried out by Frederic Kaplan\cite{Kaplan:2001}.

@@

The experiment ended with the following ``tongue-in-cheek'' post by Angus McIntyre on the Forum: \\
